NMRlipids Databank makes data-driven analysis of biomembrane properties accessible for all

### 2023-02-26 SI.tex
Update text

%\begin{abstract}
%We present a databank of lipid bilayer simulations from the NMRlipids open collaboration project.
%\end{abstract}

@@

\flushbottom
\maketitle
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
\thispagestyle{empty}

@@

@@

>>>>

\begin{abstract}
%We present a databank of lipid bilayer simulations from the NMRlipids open collaboration project.
\end{abstract}

@@

\flushbottom
\maketitle
% * <john.hammersley@gmail.com> 2015-02-09T12:07:31.197Z:
%
%  Click the title above to edit the author information and abstract
%
%\thispagestyle{empty}

@@

\begin{table}[!h]
    \centering
    %\begin{tabular}{c c c c}
    \begin{tabular}{p{3.0cm}  p{5.0cm}  p{3.0cm}  p{4.0cm}}
        Code & Function & Input & Output \\
        \hline
        \multicolumn{4}{c}{ {\bf Databank layer}}\\
        \hline
        \multicolumn{4}{c}{ {\bf Scripts/BuildDatabank/ }}\\
         AddData.py & Create README.yaml from user given information in info.yaml  & info.yaml & README.yaml \\
         searchDATABANK.py & Pair simulations with available experimental data  & README.yaml files & Updated README.yaml files \\
         QualityEvaluation.py & Quality evaluate simulations that are paired with experimental data  & README.yaml files & [lipid\_name]\_OrderParameters\_quality.json, [lipid\_name]\_FragmentQuality.json, system\_quality.json, FormFactorQuality.json \\
         \multicolumn{4}{c}{ {\bf Scripts/AnalyzeDatabank/ }}\\
         calcOrderParameters.py & Calculate C-H bond order parameters for all lipids in the simulation & README.yaml files & [lipid\_name]\_OrderParameters.json \\
         calcAPL.py & Calculate area per lipid as a function of time & README.yaml files & apl.json \\
         calc\_FormFactors.py & Calculate x-ray scattering form factors & README.yaml files & FormFactor.json \\
         calc\_thickness.py & Calculate membrane thickness & README.yaml files & thickness.json \\

    \end{tabular}
    \caption{Caption}
    \label{tab:my_label}
\end{table}

@@

\begin{table}[p]
    \centering
    \begin{tabular}{l}
    Force field name and references \\
    \hline
    CHARMM36~\cite{klauda10}\\
    Slipids~\cite{jambeck12,jambeck12b,jambeck2012another,ermilova16,grote20}  \\
    MacRog~\cite{Kulig15b}  \\
    Amber Lipid14/17~\cite{dickson14,dickson22}  \\
    Charmm-Drude~\cite{li2017drude}      \\
    ECClipids~\cite{melcr18,melcr20,bacle21}  \\
    GROMOS-CKP~\cite{Chandrasekhar03,kukol09,piggot12}  \\
    Berger~\cite{berger97}  \\
    ECC-CHARMM36~\cite{nencini2019development}  \\
    Orange~\cite{catte16}  \\
    Poger~\cite{poger10}  \\
    GROMOS 43A1-S3~\cite{chiu09}  \\ 
    GAFFlipid~\cite{dickson12}  \\
    OPLS3e~\cite{roos19}  \\
    Ulmschneider~\cite{Ulmschneider09}  \\
    Chiu Gromos~\cite{chiu09}  \\
    \end{tabular}
    \caption{List of current force fields used in simulations in the databank with references.}
    \label{tab:ForceFields}
\end{table}

### 2023-02-26 main.tex
Updates from Overleaf
Update text

\subsection{Programmatic access to MD simulations of membranes composed of most common biologically abundant lipids}

NMRlipids Databank is a community-driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve data Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} using the NMRlipids project protocol where all the content is openly accessible from the beginning of the project~\cite{botan15}.  Currently the NMRlipids databank contains nearly thousand simulation trajectories with the total length approaching half a millisecond. The distribution of available simulations containing a specific lipid in Fig.~\ref{fig:overlay}C roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}. Currently available single component lipid membranes and binary mixtures in the NMRlipids databank are illustrated in Fig.~\ref{fig:overlay}D. While these comprise the majority of available trajectories, mixtures with up to five lipid types are available. Force fields used in simulations cover all the essential parameters commonly used in lipid simulations, see Fig.~\ref{fig:overlay}E, including also united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.

The overlay structure, illustrated in Fig.~\ref{fig:overlay}A, is designed to enable efficient upcycling of MD simulations for data-driven and machine learning applications using 
The NMRlipids databank. Raw simulation data in the {\it data layer} can be stored in any publicly available location with long term stability, such as Zenodo (\url{zenodo.org}), and with permanent links to the data, such as digital object identifiers (DOIs). 
%or uniform resource names (URNs). 
The {\it Databank layer} is the core of the databank containing all the relevant information about the simulations, including links to the raw data, relevant metadata describing the systems, universal naming conventions for lipids and their atoms, computer programs to create and analyse the entries, basic properties calculated from all simulations (area per lipid, C--H bond order parameters, X-ray scattering form factors and membrane thickness), and quality evaluation of simulations against experimental data. In practise, this information is stored in the git repository which is available at~\url{https://github.com/NMRLipids/Databank}. The {\it application layer} contains outputs from the databank, such as graphical user interface (\url{http://www.databank.nmrlipids.fi/}) and results from analyses described in sections below which are available at \url{https://github.com/NMRLipids/DataBankManuscript}. A more detailed description on the NMRlipids databank structure is available the supplementary information.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay2.pdf}
    \caption{a) Schematic presentation of the overlay structure used in the NMRlipids databank. A more detailed structure of the databank layer is shown in Fig.~\ref{DatabankStructure} in the SI.
    b) Distribution of the lengths of the trajectories, total number of trajectories and total length of the simulations in the NMRlipids databank.
    c) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or fewer simulations ('others') are listed on the right. 
    d) Currently available binary mixtures in the NMRlipids databank. 
    e) Distribution of force fields in the simulations in the NMRlipids databank.
    }
    \label{fig:overlay}
\end{figure}

\subsection{Accessing NMRlipids databank}\label{section:access}
The content of the NMRlipids databank is easily searchable and browsable via graphical user interface (GUI) (\url{http://www.databank.nmrlipids.fi/}), which can be also used to visualize the systems, compare basic properties between diffent systems and experiments, and rank simulations based on their quality evaluated against experimental data.

%\subsection{Analyzing the databank}
More flexible analyses that enable wide range of novel applications can be performed by the Python application programming interface (API) available at \url{https://github.com/NMRLipids/Databank}. The API can be used to automatically analyse virtually any property from all simulations in the NMRlipids databank following the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}A. Iteration through the README.yaml files in the {\it databank layer} enables the access to raw simulation data via the links stored in these files. Affiliation of universal molecule and atom names to the simulation specific names in README.yaml and mapping files enable automatic analyses independently on the specific naming convention used in a simulation. Analysis codes and results for membrane area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness are included in the {\it Databank layer}. Further analysis codes can be implemented by users and the results can be stored to the {\it application layer} in the same format that is used in the {\it Databank layer}. The results from analyses can be then programmatically accessed and furhter analyzed using the NMRlipids databank API following the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}B.

@@

P.B. was supported by the Academy of Finland (Grants 311031 and 342908).
R.G.-F, A.P. and F.S.-L. thank Centro de Supercomputación de Galicia for computational support.
R.G.-F. thanks Ministerio de Ciencia, Innovaci{\'o}n y Universidades for a ''Ram{\'o}n y Cajal'' contract (RYC-2016-20335), and Spanish Agencia Estatal de Investigaci{\'o}n (AEI) and the ERDF (RTI2018-098795-A-I00, PDC2022-133402-I00), Xunta de Galicia and the ERDF (ED431F 2020/05, 02\_IN606D\_2022\_2667887 and Centro singular de investigaci{\'o}n de Galicia accreditation 2016-2019, ED431G/09). 
AP thanks Spanish Agencia Estatal de Investigaci{\'o}n (AEI) and the ERDF (PID2019-111327GB-I00, PDC2022-133402-I00), Xunta de Galicia and the ERDF (ED431B 2022/36, 02\_IN606D\_2022\_2667887).
F. S.-L. thanks Axencia Galega de Innovaci{\'o}n for his predoctoral contract (02\_IN606D\_2022\_2667887), and Spanish Agencia Estatal de Investigaci{\'o}n (AEI) and the ERDF (PID2019-111327GB-I00). 
M.S.M. was supported by the Trond Mohn Foundation BFS2017TMT01.
O.H.S.O, A.M.K, and R.N. acknowledge CSC − IT Center for Science for computational resources and Academy of Finland (grants 315596 and 319902) for ﬁnancial support.
\section*{Author contributions statement}
%Must include all authors, identified by initials, for example:
%A.A. conceived the experiment(s),  A.A. and B.A. conducted the experiment(s), C.A. and D.A. analysed the results.  All authors reviewed the manuscript.

>>>>

\subsection{MD simulations of membranes composed of most common biologically abundant lipids}

NMRlipids Databank is a community-driven database containing atomistic MD simulations of biologically relevant lipid membranes emerging from the NMRlipids open collaboration~\cite{botan15,ollila16,catte16,antila19,bacle21}. It has been designed to improve data Findability, Accessibility, Interoperability, and Reuse (FAIR)~\cite{wilkinson16} for MD simulation trajectories. The NMRlipids databank is constructed using the NMRlipids project protocol where all the content is openly accessible throughout the project~\cite{botan15}. Currently the NMRlipids databank contains 726 simulation trajectories with the total length of approximately 0.4\,ms. The distribution of available simulations containing a specific lipid, shown in Fig.~\ref{fig:overlay}B, roughly resembles the biological abundance of different lipid types with phosphatidylcholine (PC) being the most common followed by cholesterol, phosphatidylethanolamine (PE), phosphatidylserine (PS), phosphatidylglycerol (PG), phosphatidylinositol (PI), and other lipids depending on organism and organelle~\cite{vanmeer08}. Abbreviations and full names of all lipids present in the databank are listed in Table~\ref{tab:abbreviations}. Single component lipid membranes and binary mixtures, depicted in Fig.~\ref{fig:overlay}D, are currently most abundant in the NMRlipids databank, yet mixtures with up to five lipid types are available. Force fields used in simulations cover all the essential parameters commonly used in lipid simulations, see Fig.~\ref{fig:overlay}C and table~\ref{tab:ForceFields}, including also united atom and polarizable force fields. Therefore, the averages calculated over the databank can be considered as mean predictions from available lipid models for an average cell membrane.

The overlay structure of the NMRlipids databank, illustrated in Fig.~\ref{fig:overlay}A, is designed to enable efficient upcycling of MD simulations for data-driven and machine learning applications with mimimal investment on new infrastructure. Raw simulation data in the {\it data layer} can be stored in any publicly available location with long term stability, such as Zenodo (\url{www.zenodo.org}), and with permanent links to the data, such as digital object identifiers (DOIs). 
%or uniform resource names (URNs). 
The {\it Databank layer} (\url{www.github.com/NMRLipids/Databank}) is the core of the databank containing all the relevant information about the simulations, including links to the raw data, relevant metadata describing the systems, universal naming conventions for lipids and their atoms, quality evaluation of simulations against experimental data, and computer programs to create the entries and analyse basic properties calculated from all simulations (area per lipid, C--H bond order parameters, X-ray scattering form factors, membrane thickness and equilibration times of principal components). Also the results for these basic properties are stored in the {\it Databank layer}.

The {\it application layer} is composed of repositories and tools that read information from the {\it databank layer} for further analyses, and their results. Because the tools in the {\it application layer} do not interfere with the {\it databank layer}, they can be freely extended by anyone for wide range of purposes. Here we exemplify this with two For example, the analyses described in below sections locate in an application layer repository at \url{www.github.com/NMRLipids/DataBankManuscript}, and content of the databank can be accessed with a graphical user interface tool at~\url{www.databank.nmrlipids.fi/}. A more detailed description on the NMRlipids databank structure is available the supplementary information.

\begin{figure}
    \centering
    \includegraphics[width=\linewidth]{Figures/overlay3.pdf}
    \caption{A) Schematic presentation of the overlay structure used in the NMRlipids databank. A more detailed structure of the databank layer is shown in Fig.~\ref{DatabankStructure} in the SI.
    %b) Distribution of the lengths of the trajectories, total number of trajectories and total length of the simulations in the NMRlipids databank.
    B) Distribution of lipids present in the trajectories in the NMRlipids databank. Lipids occuring in five or fewer simulations ('others') are listed on the right. 
    C) Distribution of force fields in the simulations in the NMRlipids databank. References for each force field are given in table~\ref{tab:ForceFields}.
    D) Currently available binary mixtures in the NMRlipids databank. 
    }
    \label{fig:overlay}
\end{figure}

\subsection{NMRlipids databank-GUI: graphical access to the MD simulation data}\label{section:access}
NMRlipids databank-GUI provides easy access to the content of the NMRlipids databank via graphical user interface (GUI) that is available at~\url{www.databank.nmrlipids.fi/}. The content of the NMRlipids databank can be easily browsed and adanced searched options based on systems content and properties are available. Systems and their basic properties can be visualized and compared between different simulations and experiments. Simulations can be also ranked simulations based on their quality evaluated against experimental data.

\subsection{NMRlipids databank-API: programmatic access to the MD simulation data}
The NMRlipids databank-API, available at \url{www.github.com/NMRLipids/Databank}, enables automatic analysis of virtually any property from all simulations in the NMRlipids databank. The NMRlipids databank-API delivers functions that enable straightforward implementation of the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}A. Iteration through the README.yaml files in the {\it databank layer} enables the access to the raw data of each simulation within the databank via the links stored in these files. The affiliation of universal molecule and atom names to the simulation specific names are delivered for each simulation in README.yaml and mapping files, which enables automatic analyses over all systems independently on the specific naming convention used in a simulation. Finally, the analysis results are stored using the same structure as in the {\it databank layer}.

While analysis codes and results for basic membrane properties 
%area per lipid, C–H bond order parameters, X-ray scattering form factors and thickness 
are included in the {\it Databank layer}, unlimited amount of further analyses can be implemented by anyone in separate repositories in the {\it application layer}. {\it Application layer} repositories that are organized using the {\it Databank layer} structure can be straightforwardly accessed programmatically and fuhther analyzed by implementing the flowchart demonstrated in Fig.~\ref{fig:POPCPOPEapls}B by using the NMRlipids databank-API functions. An {\it application layer} repository used to produce the novel analyses from the NMRlipids databank exemplified in the next section is available at~\url{www.github.com/NMRLipids/DataBankManuscript}.

@@

P.B. was supported by the Academy of Finland (Grants 311031 and 342908).
R.G.-F, A.P. and F.S.-L. thank Centro de Supercomputación de Galicia for computational support.
R.G.-F. thanks Ministerio de Ciencia, Innovaci{\'o}n y Universidades for a ''Ram{\'o}n y Cajal'' contract (RYC-2016-20335), and Spanish Agencia Estatal de Investigaci{\'o}n (AEI) and the ERDF (RTI2018-098795-A-I00, PDC2022-133402-I00), Xunta de Galicia and the ERDF (ED431F 2020/05, 02\_IN606D\_2022\_2667887 and Centro singular de investigaci{\'o}n de Galicia accreditation 2016-2019, ED431G/09). 
AP thanks Spanish Agencia Estatal de Investigaci{\'o}n (AEI) and the ERDF (PID2019-111327GB-I00, PDC2022-133402-I00), Xunta de Galicia and the ERDF (ED431B 2022/36, 02\_IN606D\_2022\_2667887).
F. S.-L. thanks Axencia Galega de Innovaci{\'o}n for his predoctoral contract (02\_IN606D\_2022\_2667887), and Spanish Agencia Estatal de Investigaci{\'o}n (AEI) and the ERDF (PID2019-111327GB-I00). 
M.S.M. was supported by the Trond Mohn Foundation BFS2017TMT01.
O.H.S.O, A.M.K, and R.N. acknowledge CSC --- IT Center for Science for computational resources and Academy of Finland (grants 315596 and 319902) for financial support.
\section*{Author contributions statement}
%Must include all authors, identified by initials, for example:
%A.A. conceived the experiment(s),  A.A. and B.A. conducted the experiment(s), C.A. and D.A. analysed the results.  All authors reviewed the manuscript.
